% Copyright (C) 2013 - Michael Baudin
%
% This file must be used under the terms of the 
% Creative Commons Attribution-ShareAlike 3.0 Unported License :
% http://creativecommons.org/licenses/by-sa/3.0/

    \chapter{The Van Der Corput sequence}

\section{Example}

In this section, we present an exemple which shows how particular 
segments of the Van Der Corput behave. 
These segments are chosen so that they match the properties 
of $(t,s)$-sequences. 

Let $m,k\geq 0$ be two integers, let $b$ be 
a prime number and consider the 
elements on the Van Der Corput sequence in base $b$ 
with indices $i=kb^m,...,(k+1)b^m-1$. 
The following script presents the Van Der Corput 
sequence in base $b=2$ and considers $m=5$ and $k=13$. 
We first decompose the integer $i$ into base $b=2$. 

\lstset{language=scilabscript}
\begin{lstlisting}
b=2;
m=5;
u=lowdisc_ldgen(1000,1,"halton");
k=13;
for i=k*b^m:(k+1)*b^m-1
    d=number_tobary(i,b)';
    mprintf("i=%d=(%s)\n",i,strcat(string(d),","))
end
\end{lstlisting}

The previous script produces the following output. 
\lstset{language=scilabscript}
\begin{lstlisting}
i=416=(1,1,0,1,0,0,0,0,0)
i=417=(1,1,0,1,0,0,0,0,1)
i=418=(1,1,0,1,0,0,0,1,0)
i=419=(1,1,0,1,0,0,0,1,1)
i=420=(1,1,0,1,0,0,1,0,0)
[...]
i=443=(1,1,0,1,1,1,0,1,1)
i=444=(1,1,0,1,1,1,1,0,0)
i=445=(1,1,0,1,1,1,1,0,1)
i=446=(1,1,0,1,1,1,1,1,0)
i=447=(1,1,0,1,1,1,1,1,1)
\end{lstlisting}

We see that the 9 bits are required to decompose these 
integers into base 2. 
The four first bits are $(1,1,0,1)$, which 
is explained by the fact that $k=13=2^3+2^2+2^0$. 
The bits from 5 to 9 are all the possible bits in the set $\{0,1\}$. 

Then we decompose the i-th element of the Van Der Corput sequence 
\scivar{u(i)} into base $b=2$, printing the first 10 digits 
(the remaining low order digits are all zero).

\lstset{language=scilabscript}
\begin{lstlisting}
for i=k*b^m:(k+1)*b^m-1
    d=flps_tobary(u(i),b,10)';
    mprintf("u(i)=%f=(%s)\n",u(i),strcat(string(d),","))
end
\end{lstlisting}

The previous script produces the following output. 

\lstset{language=scilabscript}
\begin{lstlisting}
u(416)=0.021484375=(0,0,0,0,0,0,1,0,1,1)
u(417)=0.521484375=(0,1,0,0,0,0,1,0,1,1)
u(418)=0.271484375=(0,0,1,0,0,0,1,0,1,1)
u(419)=0.771484375=(0,1,1,0,0,0,1,0,1,1)
[...]
u(444)=0.240234375=(0,0,0,1,1,1,1,0,1,1)
u(445)=0.740234375=(0,1,0,1,1,1,1,0,1,1)
u(446)=0.490234375=(0,0,1,1,1,1,1,0,1,1)
u(447)=0.990234375=(0,1,1,1,1,1,1,0,1,1)
\end{lstlisting}

For example, the decomposition of 0.521484375 is 
$1/2+1/2^6+1/2^8+1/2^9$.

We see that the 4 low order bits are $(1,0,1,1)$, 
which are the reverse bits of the decomposition of $k$. 
The bits from 2 to 6 are all the possible bits in the set $\{0,1\}$. 

\section{Property}

\begin{proposition}
\label{vdc-01net}
The Van Der Corput sequence is a $(0,1)$-sequence.
\end{proposition}

\begin{proof}
Let a be an integer in the set $\{0,1,...,b^m-1\}$.
Let $I=[a/b^m,(a+1)/b^m)$ be an elementary interval. 
Its length is $1/b^m$. 
Let $k \geq 0$ be an integer. 
Let us prove that there is a unique $i$ in the set $\{kb^m,...,(k+1)b^m-1\}$  
such that $\phi(i) \in I$, where $\phi(i)$ is the i-th element 
of the Van Der Corput sequence in base $b$.
Let $a(0),a(1),...,a(m-1) \in \{0,1,...,b-1\}$ be the $m$ digits 
of the base-b decomposition of a:
\begin{eqnarray}
a=a(0) b^{m-1}+...+a(m-2) b+a(m-1).
\end{eqnarray}
Notice that the digits are ordered from the highest to the lowest 
significant digits. 
Let $a(m),a(m+1),...,a(m+r-1) \in \{0,1,...,b-1\}$ be the 
$r$ digits of the base-b decomposition of $k$ in base $b$:
\begin{eqnarray}
k=a(m+r-1)b^{r-1}+...+a(m+1)b+a(m). \label{vdc-01netdefk}
\end{eqnarray}
Notice that the digits are ordered from the lowest to the highest 
significant digits.
Let us introduce:
\begin{eqnarray}
t&=&\frac{a}{b^m} \\
 &=&\frac{a(0)}{b}+\frac{a(1)}{b^2}+...+\frac{a(m-1)}{b^m} \\
\end{eqnarray}
and
\begin{eqnarray}
x&=&\frac{a(m)}{b}+\frac{a(m+1)}{b^2}+...+\frac{a(m+r-1)}{b^r}.
\end{eqnarray}
Notice that $t$ is uniquely defined from the digits arising from the 
base-b decomposition of $a$ while $x$ is uniquely defined from the digits 
arising from the base-b decomposition of $k$. 
Consider the number 
\begin{eqnarray}
y&=&t+\frac{x}{b^m} \\
 &=&\frac{a(0)}{b}+\frac{a(1)}{b^2}+...+\frac{a(m-1)}{b^m} \nonumber \\
  &&+\frac{a(m)}{b^{m+1}}+\frac{a(m+1)}{b^{m+2}}+...+\frac{a(m+r-1)}{b^{m+r}}
  \label{vdc-01netdefy}
\end{eqnarray}
Obviously $x\geq 0$. 
Moreover, 
\begin{eqnarray}
x &\leq& \frac{b-1}{b}+\frac{b-1}{b^2}+...+\frac{b-1}{b^r} \\
  &   <& 1.
\end{eqnarray}
Indeed, this is a geometric series:
\begin{eqnarray}
\frac{b-1}{b}+\frac{b-1}{b^2}+...+\frac{b-1}{b^r}
&=& \frac{b-1}{b}(1+\frac{1}{b}+\frac{1}{b^2}+...+\frac{1}{b^{r-1}} \\
&=& \frac{b-1}{b} \frac{1-\frac{1}{b^r}}{1-\frac{1}{b}} \\
&=& \frac{b-1}{b} \frac{1-\frac{1}{b^r}}{\frac{b-1}{b}} \\
&=& 1-\frac{1}{b^r} \\
&<& 1.
\end{eqnarray}
Therefore, $x \in [0,1)$. 
Hence, $a+x \in [a,a+1)$ so that $a/b^m+x/b^m \in I$. 
But $t=\frac{a}{b^m}$, which implies $t+x/b^m=y \in I$. 
In order to conclude the proof, we recognize that $y$ 
is the i-th element of the Van Der Corput sequence in base b, 
for a particular value of $i$. 
To see this value of $i$, we just reverse the digits 
of the definition of $y$ from the equation \ref{vdc-01netdefy}. 
This leads to 
\begin{eqnarray}
i &=& a(m+r-1) b^{m+r-1}+...+a(m+1) b^{m+1}+a(m) b^m \nonumber \\
  & & a(m-1)b^{m-1}+...+a(1)b+a(0) \\
  &=& \left( a(m+r-1) b^{r-1}+...+a(m+1) b+a(m)\right) b^m \nonumber \\
  & & a(m-1)b^{m-1}+...+a(1)b+a(0) \\
  &=& kb^m+a(m-1)b^{m-1}+...+a(1)b+a(0),
\end{eqnarray}
where the last equality is implied by the equation \ref{vdc-01netdefk}.
Moreover, 
$$
a(m-1) b^{m-1}+...+a(1) b+a(0) \in \{0,...,b^m-1\}. 
$$
Hence 
$$
i \in \{b^m,...,b^m+b^m-1\}=\{b^m,...,(k+1)b^m-1\}.
$$
We have computed the unique i such that 
$y=\phi(i)$ is in I, which concludes the proof
\end{proof}
